---
title: "solutions"
author: "Edzer Pebesma"
date: "August 20, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mapview)
```

possibly after installing it. Consider the datasets `breweries`, `trails` and `franconia`, which are available after loading `mapview`. (Note that the `trails` dataset is a subset from a larger dataset, based on constraining distances to existing breweries.)

1. For each of the datasets, describe which geometry type it has, how many features (observations) and attributes (variables) it has.
```{r}
library(sf)
breweries
trails
franconia
```
2. Plot each of the three objects.
```{r}
plot(breweries)
plot(trails)
plot(franconia)
```
3. Export the `franconia` dataset to a geopackage file (`franconia.gpkg`) and try to view that file in QGIS (if you have that available on your computer).
```{r}
write_sf(franconia, "/tmp/franconia.gpkg") # may need a different path!
```
4. compare the coordinate reference systems (CRS) of the three objects. In case of differing coordinate reference systems, convert all of them to a single CRS.
```{r}
st_crs(breweries)
st_crs(franconia)
st_crs(trails)
franconia_utm = st_transform(franconia, st_crs(trails))
breweries_utm = st_transform(breweries, st_crs(trails))
```
5. For each of the trails, find out which brewery is closest to the trail.
```{r}
# with st_distance:
dists = st_distance(trails, breweries_utm)
dists[1:5, 1:5]
closests = apply(dists, 1, which.min)
closests[1:5]
# with st_nearest_feature
n = st_nearest_feature(trails, breweries_utm)
n[1:5]
# interestingly not all identical:
table(n == closests)
which(n != closests)
# check the first:
dists[181, c(closests[181], n[181])] # ambiguous question!
```
6. For each of the trails, find out the distance to the clostest brewery.
```{r}
apply(dists, 1, min)
```
7. For each of the trails, find out how many breweries it has that are less than 500 m away from the trail, per km of trail.
```{r}
apply(dists, 1, function(x) sum(x < 500))
n = apply(dists, 1, function(x) sum(x < 500))
ans = n / units::set_units(st_length(trails), km)
summary(ans)
```
8. For each of the regions in `franconia`, find out how many `breweries` it contains, and add this number as a field to the `franconia` object.
```{r}
st_contains(franconia_utm, breweries_utm)
lengths(st_contains(franconia_utm, breweries_utm))[1:5]
franconia_utm$n_breweries = lengths(st_contains(franconia_utm, breweries_utm))
```
9. For each of the regions in `franconia`, count the total `number.of.types` of beers brewed by the all the breweries in the region. 
```{r}
a = aggregate(breweries["number.of.types"], franconia, sum)
plot(a) # what could have gone wrong?
summary(breweries$number.of.types)
a = aggregate(breweries["number.of.types"], franconia, sum, na.rm = TRUE)
plot(a)
```
10. Count the total length of the trails per region in `franconia`
```{r}
i = st_intersection(franconia_utm, trails)
# wrong approach:
i$length = st_length(i)
trail_length = aggregate(i["length"], franconia_utm, sum) # TOO MANY MATCHES: touches included
# maybe better use aggregate with join = st_contains, or a different predicate?
# as another approach: join by NUTS_ID:
library(tidyverse)
j <- i %>% group_by(NUTS_ID) %>% summarize(l = sum(length))
fr_join = left_join(franconia_utm, as.data.frame(j)[c("NUTS_ID", "l")])
fr_join[1:3,]
# check:
sum(fr_join$l, na.rm = TRUE)
sum(st_length(i)) # ignores trails outside franconia
```
